---
title: "An introduction to linear mixed effects models in chronobiology: Part 2"
author: "Billy Christopher Smith"
format: html
editor: visual
---

## Load in Packages, and Setup

```{r}
#| output: false

# you may need to run: install.packages(c("dplyr","ggplot2","lme4"","MuMIn", "lmerTest", "patchwork"))

library(ggplot2) # for plotting
library(lme4) # linear mixed effects models (LMER)
library(lmerTest) # for lmer p-values
library(MuMIn) # lmer Rsq
library(patchwork) # for combining plots

okabe_ito <- c(
  "orange" =    "#E69F00",
  "sky_blue" =  "#56B4E9",
  "green"=      "#009E73",
  "yellow"=     "#F0E442",
  "blue"=       "#0072B2",
  "red"=        "#D55E00",
  "pink"=       "#CC79A7",
  "black"=      "#111111"
)

PLOTS_DIR <- "part2_figures"
dir.create(PLOTS_DIR, showWarnings = FALSE, recursive = TRUE)

```

## The data:

```{r}

df <- read.csv("data.csv", header = TRUE, sep = ",")


# -- preprocessing  --

# restrict to daytime window (drop values before 5am)
df <- df[df$pa_daybefore_maxbout_time >= 5,] 


head(df)

```

This data contains: an anonymised identifer column ***subject_id***, ***age*** and ***sex***, a day identifier ***date***, acrophase (peak time) of a diurnal rhythm extracted from wearable smartwatch heart rate (HR) data ***hr_acrophase***, and the time at which the most physically active period ('bout') of the previous day (date-1) began (***pa_daybefore_maxbout_time**)*.

The latter two variables are normalised 24-hour time-of-day variables. For example, 14.5 corresponds to 2:30pm. The variables *pa_daybefore_maxbout_time_trait* and *pa_daybefore_maxbout_time_state* can be ignored for now.

This data was derived from a wearable study of healthy volunteers under free-living (outside of the lab) conditions over 2-4 weeks. For more details, see README.md. There are 104 different subjects represented in this table. With this data, we'll ask: **“How does *when* we are most physically active affect our next-day heart rate rhythm timing?”**.

Firstly, what does the relationship between *pa_daybefore_maxbout_time* and *hr_acrophase* look like?

```{r}
# -- initial plot --
raw_plot <- ggplot(df, aes(x=pa_daybefore_maxbout_time, y=hr_acrophase))+ 
  geom_point(color="grey") # scatter points
show(raw_plot)
ggsave(paste0(PLOTS_DIR, '/', "unstandardised", ".pdf"))

```

There looks to be a small-moderate positive correlation: later activity timing seems to delay (make later) our HR rhythm timing the following day. However, people have different activity levels, schedules and exercise timing preferences: there might therefore be differences between people (identified by *subject_id)*, which a LMER is well-suited to account for.

## Back to basics: a linear regression model (*Model A)*

For comparative purposes, let's ignore the group structure for now and fit a linear model (hr_acrophase \~ pa_daybefore_maxbout_time) to this data.

```{r}
model_a <- lm(hr_acrophase ~ pa_daybefore_maxbout_time, data=df)
summary(model_a)
model_a_rsq <- summary(model_a)$adj.r.squared

LABEL_X = 14
LABEL_Y = 23

# plot model A
model_a_intercept <- model_a$coefficients[1]
model_a_slope <- model_a$coefficients[2]

model_a_plot <- ggplot(df, aes(x=pa_daybefore_maxbout_time, y=hr_acrophase))+ 
  geom_point(color="grey") + # scatter points
  geom_line(aes(y=predict(model_a)), color=okabe_ito["red"]) + # add regression line
  annotate("text", x=LABEL_X, y=LABEL_Y,
           color=okabe_ito["red"],
           label=paste0("Intercept: ", round(model_a_intercept, 2), " Slope: ", round(model_a_slope, 2), " Rsq: ", round(model_a_rsq, 2)))
ggsave(paste0(PLOTS_DIR, '/', "model_a", ".pdf"))
show(model_a_plot)
```

There is indeed a weak positive relationship (slope =0.14). The intercept here is 13.37, or 1:22pm. The model fit is weak at Rsq=0.11.

But we know this data represents 104 different subjects, what do they look like? Let's pick two of these subjects - one with an earlier than average HR acrophase, and one with a later than average HR acrophase.

```{r}
# visualise different subjects
subject_early ="e4f1d0e8bbc181058f47bdd8d21e88308ffffb722e8a2de203216c76495c2711"
subject_late = "eb03c5e0216ea3832449586327e1754b361f89e0f89435ee452024d1d798bf3c"

model_a_plot_2subj <- model_a_plot +  
  geom_point(data=df[df$subject_id==subject_early,], color=okabe_ito["blue"], size=2) + 
  geom_point(data=df[df$subject_id==subject_late,], color=okabe_ito["orange"],size=2)
ggsave(paste0(PLOTS_DIR, '/', "model_a_2subj", ".pdf"))
show(model_a_plot_2subj)

```

For the earlier subject (blue), it looks like their data follows a similar trend of a weak positive correlation. However, for the later subject (orange), there appears to be a weak negative correlation. These between-subject differences in average acrophase and activity-acrophase trend are probably contributing to the poor model fit. With this group structure, an LMER would be a better option.

## Subject-level intercept using LMERs (*Model B)*

```{r}
model_b <- lmer(hr_acrophase ~ pa_daybefore_maxbout_time + (1|subject_id), data=df)
```

Here, we've added a subject-level random intercept, by switching to lmer() and adding a (1\|subject_id) term to the formula.

```{r}
model_b_summary <- summary(model_b)
show(model_b_summary)

model_b_intercept <- model_b_summary[["coefficients"]][1, 1]
model_b_slope <- model_b_summary[["coefficients"]][2, 1]
model_b_rsq <- r.squaredGLMM(model_b)[2]

print(paste0("Intercept: ", round(model_b_intercept, 5), " Slope: ", round(model_b_slope, 5), " Rsq: ", round(model_b_rsq, 5)))
```

We get an updated estimate of the common cohort-level intercept (14.79, or 2:47pm), which takes into account the distribution of the individual subject intercepts. The common slope is even weaker at 0.04, however we have seen already that there is likely some variation in slope across the cohort. But crucially, the model fit is greatly improved, at Rsq=0.78 - much more variation in the data is now accounted for.

(A technical note: this is a specific type of R<sup>2</sup> called 'conditional R<sup>2</sup>', that takes into account the variance explained by both the model fixed and random effects. We could also look at 'marginal R<sup>2</sup>' using r.squaredGLMM(model_b)\[1\], which considers only fixed effects - you should typically report both.)

Looking at the random effects table, we see that the standard deviation of the intercept (1.802) is double the standard deviation of the residual (0.973), telling us we have more between-subject variability than within-subject.

OK, what do these subject-level intercepts look like?

```{r}

# plot with per-subject intercepts
model_b_plot <- ggplot(df, aes(x=pa_daybefore_maxbout_time, y=hr_acrophase, colour=subject_id))+ 
  geom_point() +
  # add subject level regression lines
  geom_line(aes(y=predict(model_b)), alpha=0.5) + 
  # add group-level regression line 
  geom_abline(intercept=model_b_intercept, slope=model_b_slope, linewidth=1, color=okabe_ito["red"]) +
  annotate("text", x=LABEL_X, y=LABEL_Y,
           color=okabe_ito["red"],
           label=paste0("Cohort intercept: ", round(model_b_intercept, 2), 
                        " Cohort slope: ", round(model_b_slope, 2), 
                        " Rsq: ", round(model_b_rsq, 2))) +
  # remove the legend
  theme(legend.position="none")

ggsave(paste0(PLOTS_DIR, '/', "model_b_overall", ".pdf"))
show(model_b_plot)

```

Wow - with 104 subjects, this is quite difficult to pick apart. Let's focus on our two example subjects from before.

```{r}

# highlight two subjects from before

df$model_b_predict <- predict(model_b)

# get individual intercept for each subject
model_b_coefs <- coef(model_b)$subject_id
model_b_subj_early_intercept <- model_b_coefs[subject_early, "(Intercept)"]
model_b_subj_late_intercept <- model_b_coefs[subject_late, "(Intercept)"]

model_b_plot_2subj <- ggplot(df, aes(x=pa_daybefore_maxbout_time, y=hr_acrophase))+ 
  geom_point(color="grey") +
  
  geom_point(data=df[df$subject_id==subject_early,], color=okabe_ito["blue"], size=2) +
  geom_line(data=df[df$subject_id==subject_early,], aes(y=model_b_predict), color=okabe_ito["blue"]) +
  annotate("text", x=LABEL_X, y=model_b_subj_early_intercept-2,
           color=okabe_ito["blue"],
           label=paste0("Subject intercept: ", round(model_b_subj_early_intercept, 2))) +
  
  geom_point(data=df[df$subject_id==subject_late,], color=okabe_ito["orange"], size=2) +
  geom_line(data=df[df$subject_id==subject_late,], aes(y=model_b_predict), color=okabe_ito["orange"]) +
  annotate("text", x=LABEL_X, y=model_b_subj_late_intercept+4,
           color=okabe_ito["orange"],
           label=paste0("Subject intercept: ", round(model_b_subj_late_intercept, 2))) +
  
  # add group-level regression line 
  geom_abline(intercept=model_b_intercept, slope=model_b_slope, linewidth=1, color=okabe_ito["red"]) +
  annotate("text", x=LABEL_X, y=LABEL_Y,
           color=okabe_ito["red"],
           label=paste0("Cohort intercept: ", round(model_b_intercept, 2), 
                        " Cohort slope: ", round(model_b_slope, 2),
                        " Rsq: ", round(model_b_rsq, 2))) 


ggsave(paste0(PLOTS_DIR, '/', "model_b_2subj", ".pdf"))
show(model_b_plot_2subj)

```

For the earlier subject, their individual fit looks good. For the later subject, their fit isn't terrible, but it does look like a flatter slope would be more suitable.

## Subject-level intercepts and slopes using LMERs (*Model C)*

With LMERs, we can also allow individual subject slopes to vary from the overall cohort estimate. Let's see what that looks like in terms of the model formula:

```{r}
# -- model c: introduce subject random slope --
model_c <- lmer(hr_acrophase ~ pa_daybefore_maxbout_time + (1 + pa_daybefore_maxbout_time|subject_id), data=df)
```

Here, we've changed the term "(1\|subject_id)" to "(1 + pa_daybefore_maxbout_time\|subject_id)", to specify that both the intercept (represented by 1) **and** the slope for pa_daybefore_maxbout_time are allowed to vary by subject_id.

```{r}
model_c_summary <- summary(model_c)
show(model_c_summary)

model_c_intercept <- model_c_summary[["coefficients"]][1, 1]
model_c_slope <- model_c_summary[["coefficients"]][2, 1]
model_c_rsq <- r.squaredGLMM(model_c)[2]

print(paste0("Intercept: ", round(model_c_intercept, 5), " Slope: ", round(model_c_slope, 5), " Rsq: ", round(model_c_rsq, 5)))
```

How does this model compare to our random intercepts-only model?

The intercept and slope have mainly stayed the same, and the rsq has marginally improved from 0.7762 to 0.7805. While the model is certainly a better fit, the main gain from adding random slopes is substantially reducing the degrees of freedom (df) for pa_daybefore_maxbout_time.

Let's visualise this again, for all subjects and for our two selected subjects.

```{r}
# plot with per-subject intercepts and slopes
model_c_plot <- ggplot(df, aes(x=pa_daybefore_maxbout_time, y=hr_acrophase, colour=subject_id))+ 
  geom_point() +
  # add subject level regression lines
  geom_line(aes(y=predict(model_c)), alpha=0.5) + 
  # add group-level regression line 
  geom_abline(intercept=model_c_intercept, slope=model_c_slope, linewidth=1, color=okabe_ito["red"]) +
  annotate("text", x=LABEL_X, y=LABEL_Y,
           color=okabe_ito["red"],
           label=paste0("Cohort intercept: ", round(model_c_intercept, 2), 
                        " Cohort slope: ", round(model_c_slope, 2),
                        " Rsq: ", round(model_c_rsq, 2))) +
  # remove the legend
  theme(legend.position="none")

ggsave(paste0(PLOTS_DIR, '/', "model_c_overall", ".pdf"))
show(model_c_plot)

```

```{r}
# highlight two subjects from before
df$model_c_predict <- predict(model_c)
# get individual intercept and slope for each subject
model_c_coefs <- coef(model_c)$subject_id
model_c_subj_early_intercept <- model_c_coefs[subject_early, "(Intercept)"]
model_c_subj_early_slope <- model_c_coefs[subject_early, "pa_daybefore_maxbout_time"]
model_c_subj_late_intercept <- model_c_coefs[subject_late, "(Intercept)"]
model_c_subj_late_slope <- model_c_coefs[subject_late, "pa_daybefore_maxbout_time"]

model_c_plot_2subj <- ggplot(df, aes(x=pa_daybefore_maxbout_time, y=hr_acrophase))+ 
  geom_point(color="grey") +
  
  geom_point(data=df[df$subject_id==subject_early,], color=okabe_ito["blue"], size=2) +
  geom_line(data=df[df$subject_id==subject_early,], aes(y=model_c_predict), color=okabe_ito["blue"]) +
  annotate("text", x=LABEL_X, y=model_c_subj_early_intercept-2,
           color=okabe_ito["blue"],
           label=paste0("Subject intercept: ", round(model_c_subj_early_intercept, 2), " Subject slope: ", round(model_c_subj_early_slope, 2))) +
  
  geom_point(data=df[df$subject_id==subject_late,], color=okabe_ito["orange"], size=2) +
  geom_line(data=df[df$subject_id==subject_late,], aes(y=model_c_predict), color=okabe_ito["orange"]) +
  annotate("text", x=LABEL_X, y=model_c_subj_late_intercept+4,
           color=okabe_ito["orange"],
           label=paste0("Subject intercept: ", round(model_c_subj_late_intercept, 2), " Subject slope: ", round(model_c_subj_late_slope, 2))) +
  
  # add group-level regression line 
  geom_abline(intercept=model_c_intercept, slope=model_c_slope, linewidth=1, color=okabe_ito["red"]) +
  annotate("text", x=LABEL_X, y=LABEL_Y,
           color=okabe_ito["red"],
           label=paste0("Cohort intercept: ", round(model_c_intercept, 2),
                        " Cohort slope: ", round(model_c_slope, 2),
                        " Rsq: ", round(model_c_rsq, 2)))

ggsave(paste0(PLOTS_DIR, '/', "model_c_2subj", ".pdf"))
show(model_c_plot_2subj)

```

We see that the model is able to fit a slightly flatter slope (0.01) for the later subject, which fits their data better. However, it is not negative as we might have expected from eyeballing the data. This is an example of the 'partial pooling' mentioned in Part 1 - because the trend for this subject would differ from most others, and the sample size for this specific subject is quite low, the model borrows from the overall cohort fit, and adjusts their trend accordingly.

Conversely, the slope of the earlier subject (0.03) doesn't require much adjustment from the overall cohort slope of 0.04. To bring this back to our original aim: the next-day heart rate acrophase is delayed as activity occurs later, with some variation between subjects.

What is the distribution of these slopes across all subjects?

```{r}

# but what do slopes look like overall?

model_c_slopes_mean <- mean(model_c_coefs$pa_daybefore_maxbout_time)

ggplot(model_c_coefs, aes(x=pa_daybefore_maxbout_time)) +
  geom_histogram(fill="grey", alpha=0.7) + 
  
  # mean
  geom_vline(xintercept = model_c_slopes_mean,
             colour = okabe_ito["red"], linetype = "dashed", linewidth = 1) +
  
  annotate("text", x=model_c_slopes_mean, y=15,
           color=okabe_ito["red"],
           label=paste0("Mean of subject slopes: ", round(model_c_slopes_mean, 4),
                        "\n(same as group-level slope): ", round(model_c_slope, 4))) +
  
  # early subj
  geom_segment(aes(x = model_c_subj_early_slope, y = -2, 
                   xend = model_c_subj_early_slope, yend = 0),
               arrow = arrow(),  
               size = 1,
               colour = okabe_ito["blue"]) +
  
  annotate("text", x=model_c_subj_early_slope + 0.01, y=-2.2,
           color=okabe_ito["blue"],
           label=paste0("slope=",round(model_c_subj_early_slope, 2))) +
  
  
  # late subj
  geom_segment(aes(x = model_c_subj_late_slope, y = -2, 
                   xend = model_c_subj_late_slope, yend = 0),
               arrow = arrow(),  
               size = 1,
               colour = okabe_ito["orange"]) +
  
  annotate("text", x=model_c_subj_late_slope, y=-2.2,
           color=okabe_ito["orange"],
           label=paste0("slope=",round(model_c_subj_late_slope, 2))) +
  
  
  labs(y = "Number of subjects", title="Distribution of per-subject slopes") 

ggsave(paste0(PLOTS_DIR, '/', "model_c_slopedist", ".pdf"))


```

Note that when adding random effects (intercepts and slopes), you may see singular fits / boundary convergance warnings. These often occur because too many have been added, and the random effects structure needs to be simplified.

OK, what else can we do with LMERs?

## Standardisation, and additional fixed effects (*Model D)*

In our original data table, we saw that we also have age and sex data for these subjects.

Often when performing analysis, we either want to also investigate the effects age and sex have on our response variable (hr_acrophase), or account/adjust for them so that we are observing the unique effects of our predictor of interest (pa_daybefore_maxbout_time). LMERs also allow us to easily to do both of these simultaneously by fitting **multiple fixed effects** (simply adding age and sex to the model formula). We get independent slope values for each fixed effect, which represent the relationship between that variable and the response (hr_acrophase) when the relationships between predictors have been removed.

But first, when we are fitting a model with multiple predictors, it is good practice to *standardise* them (subtract the mean, and adjust so they have a standard deviation of 1), to make it easier to directly compare their effects on the response variable.

```{r}
# when you have multiple explanatory variables, good practice to standardise 
# (mean of zero (“centering”) and standard deviation of one (“scaling”))
# so that it is easier to compare their effect sizes
for (col in c("age", "pa_daybefore_maxbout_time_trait", "pa_daybefore_maxbout_time_state", "pa_daybefore_maxbout_time")) {
  df[col] = scale(df[col], center=TRUE, scale=TRUE)
}

# what does standardised variables look like
standardised_plot <- ggplot(df, aes(x=pa_daybefore_maxbout_time, y=hr_acrophase))+ 
  geom_point(color="grey") + # scatter points
  ggtitle("Standardised")
show(standardised_plot)
ggsave(paste0(PLOTS_DIR, '/', "standardised", ".pdf"))

```

Note that pa_daybefore_maxbout_time still has the same relationship with hr_acrophase, but its values have now been adjusted.

Let's add the age and sex terms to our model.

```{r}
model_d <- lmer(hr_acrophase ~ pa_daybefore_maxbout_time + age + sex + (1+pa_daybefore_maxbout_time|subject_id), data=df)
```

```{r}
summary(model_d)

model_d_rsq <- r.squaredGLMM(model_d)[2]
print(paste0("Rsq: ", round(model_d_rsq, 5)))
```

We now have additional rows in our fixed effects table for both the age and sex effects. Crucially, we see age has a strong **negative** relationship (estimate -0.528) with hr_acrophase - as we get older, our heart rate acrophase advances.

Note that because we loaded the *lmerTest* library, estimated p-values are also displayed for each fixed effect. These tell us our estimates for pa_daybefore_maxbout_time and age are significant, however the estimate for sex is not. This doesn't mean we shouldn't include sex as an effect, however - in this case, we want to ensure we remove any sex effects, even if the effect is minimal.

## Separating lifestyle predictors into 'traits' and 'states' (*Model E)*

So it looks like pa_daybefore_maxbout_time has an effect on hr_acrophase. However, as this is free-living data, it is likely that there is substantial differences in the nature of pa_daybefore_maxbout_time between subjects: people have different work schedules and exercise timing preferences. Under the 'within-between' framework, we split such variables into the per-subject mean (their typical most-physically-active time), and the daily deviations (how much earlier or later this is than usual on a given day) from that mean. Doing so allows us to separately measure how lifestyle affects between- and within-subject variability of hr_acrophase. Borrowing terminology from psychology, we call these per-subject means 'traits', and daily deviations 'states'.

Earlier, we saw that we also have variables *pa_daybefore_maxbout_time_trait* and *pa_daybefore_maxbout_time_state* in this table. These are derived from pa_daybefore_maxbout_time, by taking the mean across the recording for a subject, and daily deviations from that mean respectively. If we added the trait and state variables back together, we would get our original pa_daybefore_maxbout_time.

Here's what these variables look like compared to our previous pa_daybefore_maxbout_time.

```{r}


# -- model e: introduce trait/state --

trait <- ggplot(df, aes(x=pa_daybefore_maxbout_time_trait, y=hr_acrophase))+ 
  geom_point(color="grey") + 
  ggtitle("Trait (subject-level average)")

state <- ggplot(df, aes(x=pa_daybefore_maxbout_time_state, y=hr_acrophase))+ 
  geom_point(color="grey") + 
  ggtitle("State (day-level deviation from average)")

trait_and_state <- standardised_plot + trait + state + plot_layout(design = "
                                                                      11
                                                                      23
                                                                   ")

show(trait_and_state)
ggsave(paste0(PLOTS_DIR, '/', "trait_and_state", ".pdf"))

```

Finally, we will fit a model with the trait and state variables as separate predictors, allowing us to separate the between- and within-subject interactions of pa_daybefore_maxbout_time on hr_acrophase. These are alongside age and sex, and a random slope for pa_daybefore_maxbout_time_state will also be added (constant variables, such as pa_daybefore_maxbout_time_trait, can't have a subject-level random slope)

```{r}
model_e <- lmer(hr_acrophase ~ pa_daybefore_maxbout_time_trait +  pa_daybefore_maxbout_time_state + age + sex + (1 + pa_daybefore_maxbout_time_state|subject_id), data=df)
summary(model_e)
```

Wow - as the estimate for the \_trait variable (1.05) is much greater than the \_state variable (0.17), this tells us the delaying effect of pa_daybefore_maxbout_time on next-day hr_acrophase is primarily driven by habitually late activity. However, the smaller estimate of the state variable is also significant - regardless of habitual activity time, later activity time still delays next-day hr_acrophase, to a smaller extent.

Furthermore, adding the trait variable has reduced between-subject variance from 2.90 to 2.02. The effect of age has been attenuated, as the trait variable now captures between-subject variance previously explained by age.

## A crucial point: circularity of data

When we first loaded in the data, we dropped rows where the maximum physical bout was below 5am, as a fix for data wrap-around.

In chronobiology, we often deal with **circular** data. For example, time of day: these values range from 0.00 to 23.99, where the value immediately after 23.99 is 0.00 - it 'wraps-around' at midnight.

**Linear** models do not work with circular data. As such, you **must** ensure that the relationship you are modelling is linear (visualisation helps), i.e no variables wrap-around. You could exclude variables as we have here, if this is reasonable for your analysis:

```{r}
df <- read.csv("data.csv", header = TRUE, sep = ",")

df$after_5am <- df$pa_daybefore_maxbout_time >= 5

ggplot(df, aes(x=pa_daybefore_maxbout_time, y=hr_acrophase, color=after_5am))+ 
  geom_point()


```

Alternatively, you could manipulate them in some way, for example we could add 24 hours to the values below 5am:

```{r}
df <- read.csv("data.csv", header = TRUE, sep = ",")

df$after_5am <- df$pa_daybefore_maxbout_time >= 5

df[df$after_5am==FALSE,"pa_daybefore_maxbout_time"] <- df[df$after_5am==FALSE,"pa_daybefore_maxbout_time"] + 24

ggplot(df, aes(x=pa_daybefore_maxbout_time, y=hr_acrophase, color=after_5am))+ 
  geom_point()


```

The trait and state separation method, based on circular mean, is another way to approach this problem.

However, depending on your data, it may not be possible to resolve these issues, in which case an LMER may not be suitable. In such cases, circular statistics or cosinor methods may work.

## Conclusion

In this document, we applied LMERs to investigate a question in real chronobiological data. We introduced random slopes, as well as how to standardise variables, and account for demographic variables using LMERs. We also introduced the concept of 'traits' and 'states' to distinguish between- and within-subject effects of lifestyle on chronobiology.
