---
title: "An introduction to linear mixed effects models in chronobiology: Part 1"
author: "Billy Christopher Smith"
format: html
editor: visual
---

## What is this document?

This is a 'Quarto' document. Quarto enables you to put content and executable R code together into a single document. Quarto .html files are rendered from .qmd files, which are similar to Jupyter Notebooks. If you open a .qmd in RStudio, you can make changes to the code, and press 'Render' to update outputs and convert this into a .html (web page) document. To learn more about Quarto, see <https://quarto.org>.

## Hello, world!

Below is some simple R code. We store some text in a variable 'my_string', and output this with a friendly greeting.

When we run this, quarto pastes the output below the code. You would see the same output if you ran this code directly in RStudio.

```{r}
my_string <- "Autumn Clock Club 2026"
print(paste0("Hello, ", my_string, "!"))

```

## Load in Packages, and Setup

```{r}
#| output: false

# you may need to run: install.packages(c("dplyr","ggplot2","lme4","bayestestR","MuMIn"))

library(dplyr) # many functions for handling data
library(ggplot2) # for plotting
library(lme4) # linear mixed effects models (LMER)
library(bayestestR) # for generating simpson's data
library(MuMIn) # for r.squaredGLMM

okabe_ito <- c(
  "orange" =    "#E69F00",
  "sky_blue" =  "#56B4E9",
  "green"=      "#009E73",
  "yellow"=     "#F0E442",
  "blue"=       "#0072B2",
  "red"=        "#D55E00",
  "pink"=       "#CC79A7",
  "black"=      "#111111"
)

PLOTS_DIR <- "part1_figures"
dir.create(PLOTS_DIR, showWarnings = FALSE, recursive = TRUE)



```

## Here's some data - what's the relationship?

```{r}

set.seed(2026) # set RNG seed for simpson's data generation

df <- simulate_simpson(
    n = 30,
    r = -0.6,
    groups = 8,
    difference = 1,
    group_prefix = "G_"
  )

names(df) <- c("X", "Y", "Group")
```

Here, we are generating some data for this tutorial. The options (parameters) we're passing into this method here allow us to specify exactly the relationship present in this data. Simulated data where we know the exact relationship present in the data is usually a good way to test our understanding when applying a new type of statistical model.

Typically, R data takes the form of 'data.frames' (df): a representation of data in column and row form, much like a spreadsheet.

When working with new data, whether it is something we've generated or loaded in from a .csv file, we should inspect the structure of the data, and visualise relationships between variables in the table:

```{r}

# What does the table look like?
head(df)

# What does the relationship look like? (scatter plot of the X and Y variables)
ggplot(df, aes(x=X, y=Y))+ 
  geom_point() 

```

```{r}
#| output: false
#| 
# Save this plot
ggsave(paste0(PLOTS_DIR, '/', "scatter", ".pdf"))
```

Interesting - it looks like there is a positive linear correlation between the two variables - as X increases, Y also generally increases.

However, we missed something when we looked at the raw data above - this data actually contains samples from 8 different **groups**. In general, groups in data could be subjects, patients, sites, cells, plants, animals, etc. We should check whether this relationship also holds for the groups within.

```{r}

# Scatter plot, colour by group
ggplot(df, aes(x=X, y=Y, color=Group))+ 
  geom_point() + scale_colour_manual(values = unname(okabe_ito))
```

```{r}
#| output: false
ggsave(paste0(PLOTS_DIR, '/', "scatter_with_groups", ".pdf"))
```

Wow - it turns out these groups actually each have a **negative** linear correlation - as X increases, Y **decreases**. When viewed together, overlap between groups makes the overall relationship appear positive.

This is an example of [Simpson's Paradox](https://en.wikipedia.org/wiki/Simpson%27s_paradox) - where the trends in our data change substantially when we consider its **group** structure. *Linear mixed effects models* can help us account for this, and we'll see why that is in the following sections.

## But my data doesn't look like 'Simpson's paradox': why do I need LMERs?

Yes, you don't need to have this exact situation in your data to use LMERs.

You might want to use LMERs whenever you have **group structure** in your data. For example, if you wanted to capture the rate of evening melatonin increase over 6 hourly samples for each of 10 study participants, or change in clock gene periodicity over daily samples for plants at 10 different sites. In both cases, each sample is not independent, but related to other samples within the same group. As such, it's important and beneficial to model that there are group differences, and helps avoid the issue of [pseudoreplication](https://scienceinsights.org/what-is-pseudoreplication-in-experimental-design/).

## Linear regression models, intercepts, slopes, and R<sup>2</sup>

Let's make a simple linear regression model for our Simpson's paradox data.

```{r}
simpsons_lm <- lm(Y ~ X, data=df)
```

Here, we are setting up a linear regression model (*lm*) where we define that Y (our response variable) is some function of X (our predictor variable). For example, at any given time, *Y* might be double the value of *X* (Y = 2X): if X was 3, Y would be 6, and so on. We would call this a positive linear correlation, with a *slope* of 2.

Alternatively, if Y=-2X, we would call this a negative linear correlation with a slope of -2. Fitting a model helps you measure the relationship between your variables. It finds a number for important values such as the slope.

Finally, we store our model inside a variable called '*simpsons_lm*' - this isn't strictly necessary, but is helpful in case we want to look at our model again later.

OK, let's look at the output of this model:

```{r}
summary(simpsons_lm)
```

There's a lot of information here, the most important being the **intercept** and **slope**. We've already covered the slope. The intercept is simply the value of Y when X is 0, which offsets the trend (e.g. Y = 2X + Slope). You can find the intercept and slope in the table above, under the first column ('Estimate') of the table underneath 'Coefficients:'.

```{r}
simpsons_lm_intercept <- simpsons_lm$coefficients[1]
simpsons_lm_slope <- simpsons_lm$coefficients[2]

print(paste0("Intercept: ", round(simpsons_lm_intercept, 5), " Slope: ", round(simpsons_lm_slope, 5)))

```

To see what these represent, let's visualise the data and the regression line fit by our model:

```{r}
simpsons_lm_rsq <- summary(simpsons_lm)$adj.r.squared

simpsons_lm_plot <- ggplot(df, aes(x=X, y=Y))+ 
  geom_point(color="grey") + # scatter points
  geom_line(aes(y=predict(simpsons_lm)), color=okabe_ito["red"]) + # add regression line
  annotate("text", x=5, y=10,
           color=okabe_ito["red"],
           label=paste0("Intercept: ", round(simpsons_lm_intercept, 2), " Slope: ", round(simpsons_lm_slope, 2), " Rsq: ", round(simpsons_lm_rsq, 2)))
show(simpsons_lm_plot)
```

```{r}
#| output: false

ggsave(paste0(PLOTS_DIR, '/', "simpsons_lm", ".pdf"))
```

The model **slope** is 0.75 - as X increases by 1, Y increases by 0.75 (a positive linear correlation).

The **intercept** - the value of Y when X is 0 - is 1.12.

To best fit our regression line to the data, the model intercept helps us place our line vertically, and the model slope helps us rotate it. Using our previous notation, this regression line is produced with the formula Y=0.75X + 1.12.

Not previously mentioned: the model R<sup>2</sup> (***Rsq***) is 0.56. R<sup>2</sup> is a value that ranges between 0 and 1, that tells us how 'good' the fit intercept and slope are at describing the relationship in the data. A value of 1 represents an (unrealistically) perfect fit, where all the points fall on the regression line. A value of 0 means that none of the variability in the data is captured by the model. In reality, values of 0 or 1 are very rare - most fall somewhere in between. 0.56 represents a moderate fit: however we know that this model is mistaken, as it doesn't consider the group structure. To learn more about R<sup>2</sup>, see [Coefficient of determination](https://en.wikipedia.org/wiki/Coefficient_of_determination).

However, we know that this is not the true nature of the data. Let's see how fitting a *linear mixed effect regression model* (LMER) differs.

## Linear mixed effect regression model (LMER)

To fit an LMER, we use a library called 'lme4'. Libraries extend what we can do with R code, by adding extra functionality.

```{r}
simpsons_lmer <- lmer(Y ~ X + (1|Group), data=df)
```

Just as before, we are setting up a model where we define that Y (our response variable) is some function of X (our predictor variable). In LMER terminology, our predictor X is a '**fixed effect**'. It's something that we assume has a consistent 'effect' on the response variable across any groups we have within the data.

In our visualisation above, we saw that each group in this dataset had a similar negative correlation (i.e, the same slope), but were slightly offset from each other. In an LMER model, we can capture these different group-level offsets using a **random effect**. This is what the new term '*(1\|Group)*' is doing: the intercept (denoted by 1) varies by/is dependent upon the 'Group' variable. Even if we aren't primarily interested in the difference between groups, adding this random effect ensures we are modelling the data properly. The grouping variable must be categorical (a factor), and a good rule-of-thumb is to have at least 5 different groups in your data (e.g 5 subjects or sampling sites), though variance estimates can be unreliable below 10-20 groups.

Note that LMERs are also referred to as LMMs, mixed-effects models, multilevel models, and hierarchical models.

Before we look at the LMER output table, let's first visualise the LMER model:

```{r}
simpsons_lmer_summary <- summary(simpsons_lmer)

# group-level intercepts and slope
simpsons_lmer_intercept <- simpsons_lmer_summary[["coefficients"]][1, 1]
simpsons_lmer_slope <- simpsons_lmer_summary[["coefficients"]][2, 1] 

# overall model Rsq
simpsons_lmer_rsq <- r.squaredGLMM(simpsons_lmer)[2]


# plot with per-subject intercepts
simpsons_lmer_plot <- ggplot(df, aes(x=X, y=Y, colour=Group))+ 
  geom_point(alpha=0.25) +
  # add participant level regression lines
  geom_line(aes(y=predict(simpsons_lmer))) + scale_colour_manual(values = unname(okabe_ito)) + 
  # add group-level regression line
  geom_abline(intercept=simpsons_lmer_intercept, slope=simpsons_lmer_slope, linewidth=1, color=okabe_ito["red"], linetype="dashed") +
  annotate("text",  x=2, y=9,
           color=okabe_ito["red"],
           label=paste0("Intercept: ", round(simpsons_lmer_intercept, 2),
                        " Slope: ", round(simpsons_lmer_slope, 2), 
                        " Rsq: ", round(simpsons_lmer_rsq, 2)))


show(simpsons_lmer_plot)
```

```{r}
#| output: false
ggsave(paste0(PLOTS_DIR, '/', "simpsons_lmer", ".pdf"))
```

This looks completely different!

Each group has its own regression line. They have the same slope, which is now correctly negative. Furthermore, as we specified that the intercept can vary, the regression lines can have different vertical positions, which allows better fit to group data. Consequently, the overall model fit (Rsq=0.96) has drastically improved.

At this point, we might ask: **why use LMERs for this? Why not just run 8 separate linear regressions, one for each group?** Well, an additional benefit of LMERs is that they implement 'partial pooling'. We assume the groups are a sample from a larger population (for example, 8 people selected at random: note that the structure of the simpson's paradox data is quite unrealistic - we'll see more realistic data in part 2. ). Rather than generate 8 completely unique intercepts (8 separate regressions, or 'no pooling'), we estimate a common intercept, and model the random intercepts as deviations from this common value. This has the added benefit of using information from groups with more samples to improve the model estimates for groups with fewer samples.

As such, we get an intercept and slope that represent the common trend across all groups (the dashed red regression line, with intercept 7.15 and slope -0.59), as well as the intercepts for individual groups (which are visualised as regression lines above):

```{r}
coef(simpsons_lmer)
```

**As an exercise: fit separate linear models for each group, and compare their intercepts to those provided by the LMER.**

Note that, in this data, only the intercept varies between groups. It is very likely that the slope may vary between groups too, which LMERs can also handle - we'll cover that in Part 2.

As before, we can also find these values in the tabular output:

```{r}

show(simpsons_lmer_summary)
```

Like the table under 'Coefficients:' for the linear regression model, we find a table with the model intercept, and slope for the X variable, under 'Fixed effects:'.

```{r}
print(paste0("Intercept: ", round(simpsons_lmer_intercept, 4), " Slope: ", round(simpsons_lmer_slope, 4), " Rsq: ", round(simpsons_lmer_rsq, 2)))
```

Under 'Random effects:', we find useful information on the how well the model captures variance between our groups. This tells us that there is a variance between groups of 15.1217 with a standard deviation of 3.8887, and a residual (unexplained) variance of 0.6429 with standard deviation of 0.8018. As the group standard deviation is larger than the residual, this tells us most variability is between groups, rather than within.

## A side note: 'long' and 'wide' data?

It should be noted that your data table usually needs to be in 'long' format to work with plotting functions and LMERs in R.

You can check this by looking at the data table structure:

```{r}
# What does the table look like?
head(df)
```

As the identifying column (Group) is repeated, this is referred to as 'long' format, where each row represents a sample.

If you instead had 1 row per group, this would be 'wide' format.

If your data is in wide format, not to worry - there [ are methods in R](https://www.geeksforgeeks.org/r-language/reshaping-data-frame-from-wide-to-long-format-in-r/) (and other languages) to help you 'reshape' your data from one format to the other.

You can learn more about wide and long formats [here](https://www.statology.org/long-vs-wide-data/).

## Conclusion

In this document, we looked at the importance of capturing group-level differences in data, and how to apply both linear regression models and linear mixed effects regression models in the R programming language.

Now that we have seen how LMERs work on some toy data, let's apply them to answer questions in real chronobiological data.
